`flippant` -- An `R` package for the automated  analysis of dithionite

### 2015-09-16 inst/manuscript/manuscript.rmd
Provide proper citing facilities and start to fill in the citations; also add some clarification comments to ask Birgit;

---
title: "`flippant` -- An `R` package for the automated  analysis of dithionite 
  scramblase assays"
author: "Johannes Graumann ^†,^*, Birgit Ploier ^‡^, Richard J. Cotton ^†^,
  Mike A. Goren ^‡^ and Anant K. Menon ^‡^"
date: "September 16, 2015"
output: 
#  word_document
  pdf_document:
    fig_caption: yes
#bibliography: flippant.bib
#csl: bioinformatics.csl
---

@@

# Introduction
Scramblases are proteins mediating the mass-action driven equilibration of 
<CITE SOME REVIEW>. Indipendent of energy equivalents such as ATP they 
lipids between the leaflets of lipid bilayers representing biological membranes
facilitate this ``flipping'' by providing a mechanism of transition through the 
hydrophobic environment of membranes to polar lipid head groups, thus 
reducing the massive energetic cost that process would carry otherwise <CITE 
ANANT HALF TIME>. While the underlying molecular mechanisms remain enigmatic 
<CITE SOMETHING>, and molecules supporting flipping processes as central as the 
equilibration of the ubiquitous phosphoglycerolipids across the membrane of the 
endoplasmic reticulum, where they are synthesized in the cytoplasmic leaflet, 
have yet to be identified <CITE SOMETHING>, there has been much recent progress 
in scramblase identification and characterization. Molecules shown to have 
scramblase activity include opsin/rhodopsin <CITE INDU, MIKE, BIRGIT>, X <CITE> 
and Y <CITE>.

Following work by X <CITE>, Y<CITE> and Z<CITE>, we have develloped an assay 
suited to characterize scramblase activity in purified proteins and protein 
extracts <CITE OPSIN PAPER, MIKE, BIRGIT>, based on the reconstitution of 
scramblases or candidate proteins into large synthetic vesicles with a lipid
composition that includes the fluorescent lipid XYC. Upon exposure of these 
vesicles to membrane-impermeable reducing agents such as dithionite, 
fluorophores in the external sheet of the vesicles are quenched and fluorescence
in scramblase-free vesicles accordingly quickly drops to approximately 50% of 
the initial value. However, for vesicles reconstituted with proteins conferring 
a scramblase activity, fluorescence dissapears entirely, as fast 
scramblase-mediated equilibration between outer (exposed) and inner (protected)
sheet of the vesicle membrane renders all fluorescent lipids accessible to 
reduction. By titration of the protein amount reconstituted into the vesicle 
membranes it is thus possible to characterize scramblases and probe their 
activity. Beyond functional validation of candidate scramblases 
<CITE OPSIN, Birgit>, the assay has for example also been used to characterize 
the stoichiometry of scramblase homo-multimers reconstituting into the vesicles
<CITE BIRGIT>.

From the identification of baseline and post-reduction fluorescense from 
fluorometer-generated spectra, via the calculation of the underlying statistics
and through publication-grade representation of the results, data analysis in 
the context of this assay has until now involved an extensive series of manual 
steps, at most supported by spread sheet calculation facilities. Based on the
`R` free software environment for statistical computing and graphics <CITE CORE>
and prominently drawing on the power and elegance of the `ggplot2` plotting 
system <CITE the book>, we here provide `flippant` a comprehensive package
covering fast, reproducible data analysis and graphing for scramblase assays 
from raw fluorescense spectra with a simple parametrizing spreadsheet as the
sole required input.

@@

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

>>>>

---
title: "`flippant` -- An `R` package for the automated  analysis of dithionite 
  scramblase assays"
author: "Johannes Graumann ^†,^*, Birgit Ploier ^‡^, Richard J. Cotton ^†^,
  Mike A. Goren ^‡^ and Anant K. Menon ^‡^"
date: "September 16, 2015"
output: 
#  word_document
  pdf_document:
    fig_caption: yes
bibliography: flippant.bib
csl: bioinformatics.csl
---

@@

# Introduction
Scramblases are proteins mediating the mass-action driven equilibration of 
lipids between the leaflets of lipid bilayers representing biological membranes
[@waechter_flippases_2013]. Indipendent of energy equivalents such as ATP they 
facilitate this ``flipping'' by providing a mechanism of transition through the 
hydrophobic environment of membranes to polar lipid head groups, thus 
reducing the massive energetic cost that process would carry otherwise <CITE 
ANANT HALF TIME>. While the underlying molecular mechanisms remain enigmatic 
<CITE SOMETHING>, and molecules supporting flipping processes as central as the 
equilibration of the ubiquitous phosphoglycerolipids across the membrane of the 
endoplasmic reticulum, where they are synthesized in the cytoplasmic leaflet, 
have yet to be identified <CITE SOMETHING>, there has been much recent progress 
in scramblase identification and characterization. Molecules shown to have 
scramblase activity include opsin/rhodopsin 
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. **Would be good to include/replace some 
with non-in-house identifications/characterizations. What else is out there?**

To interrogate factors of this functional class, we have developed a biochemical 
assay suited to characterize scramblase activity in purified proteins and protein 
extracts [@chang_chemical_2004], based on the reconstitution of scramblases or 
candidate proteins into large synthetic vesicles with a lipid
composition that includes nitrobenzoxadiazole (NBD) labeled fluorescent lipids.
Upon exposure of these vesicles to membrane-impermeable reducing agents such as 
dithionite, fluorophores in the external leaflet of the vesicles are quenched and 
fluorescence in scramblase-free vesicles accordingly quickly drops to 
approximately 50% of the initial value. However, for vesicles reconstituted with
proteins conferring a scramblase activity, fluorescence dissapears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the vesicle membrane renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the vesicle membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@goren_constitutive_2014;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014],
the assay has for example 
also been used to characterize the stoichiometry of scramblase homo-multimers 
reconstituting into the vesicles [@ploier_perturbed_submitted] and probe the 
ATP-dependence of flippases, molecules that under energy-consumption flip lipids
against a concentration gradient between bilayer leaflets [@soupene_atp8a1_2008].
**Are you aware of further use? More adoption would fortify the need for the package ...**

From the identification of baseline and post-reduction fluorescense from 
fluorometer-generated spectra, via the calculation of the underlying statistics
and through publication-grade representation of the results, data analysis in 
the context of this assay has until now involved an extensive series of manual 
steps, at most supported by spreadsheet calculation facilities [see the 
supplementary material of @menon_opsin_2011 for extensive stepwise discussion]. Based on the
`R` free software environment for statistical computing and graphics [@r_core_team_r:_2015]
and prominently drawing on the power and elegance of the `ggplot2` plotting 
system [@wickham_ggplot2_2009], we here provide `flippant`, a comprehensive package
covering fast, reproducible data analysis and graphing for scramblase assays 
from raw fluorescense spectra with a simple parametrizing spreadsheet as the
sole required input.

@@

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
#References
